WNBA vs NBA Earnings In Comparable Players

### 2022-12-05 Manuscript/Manuscript.tex
Finalized Manuscript

\begin{abstract}
In a world where gender pay gaps have been highlighted in nearly every industry, professional sports are no exception. The Women’s National Basketball 
Association is widely known and accepted to pay very significantly less annually to its players than the male league counterpart, the National Basketball 
Association. There has been a lot of research into this pay gap, and this paper will look to add to the conversation by examining the structure of 
contracts in relation to performance. This will be done using statistics from the 2018 seasons of each respective league, using datasets cleaned up and 
analyzed with various python scripts to eventually draw conclusions on the relation between performance and pay for male and female athletes.	
\end{abstract}

@@

@@

gender roles and cultural norms of the time, however, women at the time were believed to be very fragile and unable to play physical sports like 

@@

end and defend the other team on the other end. In 1892, however, women initially played “9 on 9”, with the court divided into three areas: offense, 
defense, and mid-court. Each player was assigned one of these courts in which they played, such that 3 women on each team were in each section. The 

@@

attributed to the fear that the women players might suffer “nervous fatigue” if the game were too strenuous \citep{WNBA_Hist}.

@@

beginning in the late 19th century, but did not enter the professional realm until the 1970s. The sport became officially recognized at the collegiate 

@@

There are minimal differences between today's women’s basketball and the basketball that men have been playing since inception. The WNBA and wome's 

@@

examine the correlation between individual player performance and performance in numerous advanced statistics. The goal is to draw conclusions on if there 

@@

\section{Data}
There are two primary data sets that I will be using to conduct the analysis in this paper. The first of these is a database of historical WNBA data 
dating from 2001-2020, collected by Neil Payne, sports editor at 538, and displayed on his GitHub account \citep{first}. Unlike most data sets, this one 
contains a multitude of advanced statistics for each player.

@@

\section{Research Design and Methods}
Initially, we began by performing a series of paired sample t-tests to confirm some expectations from the data. We do not need to perform a test to 
determine if NBA players are paid more than WNBA players, as that is commonly known and very obvious. As $\textit{Forbes}$ basketball author David Berri 
discovered, in general WNBA players are still paid significantly less than NBA players proportional to league revenue \citep{WNBA_Revenue}. In order to 
confirm that the data for the 2018 seasons fit with this mold, an initial t-test was performed:
$$H_0 = \mu_1 \leq \mu_0, H_a = \mu_1 > \mu_0$$
where $\mu_0$ represents the average WNBA salary proportional to league revenue and $\mu_a$ represents the same for the NBA. The test was done using a 
significance level of  5\%, or an alpha value of 0.05, and degrees of freedom $n_1 + n_2 - 2 = 516$. By the t-statistic chart, it must be checked if the 
found t-statistic is greater than 1.646. Using a Python script, Comp\_t\_tests.py, it was found that our t-statistic was 16.56. Since we want to convert 
this into a one tailed test, we halve the t-statistic to get 8.28, which is far greater than 1.646, so it can safely be said with 95\% confidence that our 
data follows the trend found in Berri's article.
\par
The primary focus of the analysis section was to analyze correlation using linear regression between advanced statistics and salary. This was done once 
again using python scripts: LR\_Test\_WNBA and LR\_Test\_NBA, for the WNBA and NBA, respectively, were used to read the data from the cleaned datasets and 
conduct simple linear regression analysis on all the advanced stats against player salaries. Then, using the scipy and matplotlib Python packages, each 
comparison was graphed with a line of best fit and a correlation coefficient ($R^2$) was found. This correlation coefficient tells us how strong the 
correlation effect is between the statistic and salary, which will help us draw significantly more accurate conclusions from the data then if simply the 
slope was examined. There were twelve overlapping statistics between the WNBA and NBA datasets, and the correlation coefficients for these statistics in 
both leagues are graphed below, with the standard error in parenthesis:

@@

 PER  &  0.019458 (1.576)  & 0.25444 (0.014) \\
 FTr   &  0.001641 (0.036)  & 0.033096 (0.009)       \\ 
 ORB\%   &  0.019717 (0.929)  & 0.000274 (0.273)       \\   
 TRB\%   &  0.000017 (1.080)   & 0.007801 (0.337)       \\
 AST\%   &  0.041707 (1.936)  & 0.110984 (0.608)    \\
 STL\%  &  0.003666 (0.175)  & 0.006126 (0.045)       \\    
 BLK\%  & 0.000444 (0.337)   & 0.000141 (0.114)     \\
 TOV\%  & 0.010470 (1.432)  & 0.004207 (0.315)    \\
 USG\%   & 0.033326 (1.146)   & 0.202527 (0.350)       \\
 OWS   & 0.117693 (0.276)   & 0.337278 (0.121)        \\
 DWS  & 0.151183 (0.133)   & 0.234252 (0.068)         \\
 WS   &  0.153999 (0.365)  & 0.364165 (0.164)                 \\
 (WS/40)/(WS/48)   &  0.047009 (0.022)  & 0.151640 (0.004) \\
 \hline
\end{tabular}
\newline
\par
The graphs for the strongest correlation in both leagues (both the win shares statistic) are shown below:
\newline
\includegraphics[width=.85\textwidth]{NBA_WS_Graph.png}
\newline
\includegraphics[width=.85\textwidth]{WNBA_WS_Graph.png}

\par
Finally, within the Comp\_t\_tests.py Python script, analysis was conducted on the salaries of the top players in each league. Instead of looking at 
simple statistics, work that has been done in (DiCicio), this method used the advanced statistics from the data set. Six statistics were selected: Minutes 
Played (MP), Player Efficiency Rating (PER), Free Throw rate (FTr), Offensive Win Shares (OWS), Defensive Win Shares (DWS), and Win Shares (WS). The 
script was written to sort the datasets by these statistics, and the top 10 players in each league in these statistics were extracted. Then the salaries 
were compared to each respective league's median salary, and the average of these top 10 players relations to the average were compiled. The entries in 
the graph can be interpreted as follows: If the value is greater than 1, than the top 10 players in that statistic are paid the value as a factor above 
league average. If the value is less than 1, than the top 10 players in that statistic are paid the value as a factor below league average. The graph is 
below:
\vspace*{1cm}
\begin{tabular}{*5l} 
\toprule
How Much Compared to Avg & the Top 10 Players Are Paid & For Advanced Stats \\\midrule

@@

The data found from the initial t-tests supported what had been found in the current literature. It was confirmed that in our 2018 data WNBA players are 
paid significantly less than NBA players, a known fact. It was also confirmed with a very high degree of certainty that WNBA players are paid more than 
NBA players proportional to league revenue (around 60 million for the WNBA, 8 billion for the NBA). This confirms the findings of (DiCicio), but was found 
to be particularly interesting and is in general counterintuitive to the narrative surrounding pay in women’s basketball. It also changes the conversation 
on how the pay gap can be narrowed.

@@

when compared to Win Shared is relatively high, the NBA graph looks much more traditionally distributed: Examining it finds the data generally following 

@@

\section{Conclusion}
The motivation of this paper was to try and determine what goes into the pay gap in men’s and women’s basketball and primarily to see if the salary 
models were similar (just dramatically scaled down in the much poorer league). It was confirmed that the data for the 2018 season fit with precedent 
in terms of NBA salaries being much higher, both base and proportional to league revenue. The regression analysis revealed that the correlation 
coefficients for NBA salaries to selected impactful advanced statistics was in general much higher than that of the WNBA, revealing that the 
aforementioned salary models were dramatically different. It was, however, determined that finishing in the top 10 in said statistics resulted in pay 
increases over league median, however the NBA’s pay increases were significantly more due to their much larger spread in salaries. 
\par
Perhaps one of the most interesting conclusions is drawn when examining the graphs of NBA and WNBA pay distribution against Win Shares performance. 
Not only is there much less pay increase opportunity in the WNBA, but it seems as though almost all players are grouped into one of two pay rates, 
which may or may not be based solely upon performance.
\par
Next steps from this data involve researching what goes into league revenue and attempting to determine why it has never been profitable. 
It is important to acknowledge one last time that the WNBA is hamstrung by this lack of revenue, but the contract structures and much lower 
pay proportional to league revenue reinforce male dominant stereotypes in an already masculine world with a widespread gender pay gap in nearly
every industry. Ultimately, the NBA players get rewarded much more for performance than the WNBA, which may motivate them to improve more and bring 
more excitepment and popularity to the league. With less monetary motivation to perform, the WNBA continues to sell itself well short of its potential 
and does not give the players the respect that they deserve.

>>>>

\begin{abstract}
In a world where gender pay gaps have been highlighted in nearly every industry, professional sports are no exception. The Women’s National Basketball 
Association is widely known and accepted to pay very significantly less annually to its players than the male league counterpart, the National Basketball 
Association. There has been a lot of research into this pay gap, and this paper will look to add to the conversation by examining the structure of 
contracts in relation to performance. This will be done using statistics from the 2018 seasons of each respective league, using datasets cleaned up and 
analyzed with various Python scripts to eventually draw conclusions on the relation between performance and pay for male and female athletes.	
\end{abstract}

@@

\hspace*{10mm}

@@

gender roles and cultural norms of the time, however, women were believed to be very fragile and unable to play physical sports like 

@@

end and defend the other team on the other. In 1892, however, women initially played “9 on 9”, with the court divided into three areas: offense, 
defense, and mid-court. Each player was assigned to play within one of these cours, such that 3 women on each team were in each section. The 

@@

attributed to the fear that the players might suffer “nervous fatigue” if the game were too strenuous \citep{WNBA_Hist}.

@@

beginning in the late 19th century, but did not enter the professional realm until the 1970s. Women's basketball became officially recognized at the collegiate 

@@

There are minimal differences between today's women’s basketball and the basketball that men have been playing since inception. The WNBA and women's 

@@

examine the correlation between individual player pay and performance in numerous advanced statistics. The goal is to draw conclusions on if there 

@@

\section{Data}
\hspace*{10mm}
There are two primary data sets that I will be using to conduct the analysis in this paper. The first of these is a database of historical WNBA data 
dating from 2001-2020, collected by Neil Payne, sports editor at 538, and displayed on his GitHub account \citep{first}. Unlike most data sets, this one 
contains a multitude of advanced statistics for each player.

@@

\section{Research Design and Methods}
\hspace*{10mm}
Initially, we began by performing a paired sample t-test to confirm some expectations from the data. We do not need to perform a test to 
determine if NBA players are paid more than WNBA players, as that is commonly known and very obvious. As $\textit{Forbes}$ basketball author David Berri 
discovered, in general WNBA players are still paid significantly less than NBA players proportional to league revenue \citep{WNBA_Revenue}. In order to 
confirm that the data for the 2018 seasons fit with this mold, an initial t-test was performed:
$$H_0 = \mu_1 \leq \mu_0, H_a = \mu_1 > \mu_0$$
where $\mu_0$ represents the average WNBA salary proportional to league revenue and $\mu_a$ represents the same for the NBA. The test was done using a 
significance level of  5\%, or an alpha value of 0.05, and degrees of freedom $n_1 + n_2 - 2 = 516$. By the t-statistic chart, it must be checked if the 
found t-statistic is greater than 1.646. Using a Python script, Comp\_t\_tests.py, it was found that our t-statistic was 16.56. Since we want to convert 
this into a one tailed test, we halve the t-statistic to get 8.28, which is far greater than 1.646, so it can safely be said with 95\% confidence that our 
data follows the trend found in Berri's article.
\par
The primary focus of the analysis section was to analyze correlation using linear regression between advanced statistics and salary. This was done once 
again using python scripts: LR\_Test\_WNBA and LR\_Test\_NBA, for the WNBA and NBA, respectively, were used to read the data from the cleaned datasets and 
conduct simple linear regression analysis on all the advanced stats against player salaries. Then, using the scipy and matplotlib Python packages, each 
comparison was graphed with a line of best fit and a correlation coefficient ($R^2$) was found. This correlation coefficient tells us how strong the 
correlation effect is between the statistic and salary, which will help us draw significantly more accurate conclusions from the data then if simply the 
slope was examined. There were twelve overlapping statistics between the WNBA and NBA datasets, and the correlation coefficients for these statistics in 
both leagues are shown below, with the standard error in parenthesis:

@@

 PER  &  0.019458 (1.576)  & 0.25444 (0.014) \\
 FTr   &  0.001641 (0.036)  & 0.033096 (0.009)       \\ 
 ORB\%   &  0.019717 (0.929)  & 0.000274 (0.273)       \\   
 TRB\%   &  0.000017 (1.080)   & 0.007801 (0.337)       \\
 AST\%   &  0.041707 (1.936)  & 0.110984 (0.608)    \\
 STL\%  &  0.003666 (0.175)  & 0.006126 (0.045)       \\    
 BLK\%  & 0.000444 (0.337)   & 0.000141 (0.114)     \\
 TOV\%  & 0.010470 (1.432)  & 0.004207 (0.315)    \\
 USG\%   & 0.033326 (1.146)   & 0.202527 (0.350)       \\
 OWS   & 0.117693 (0.276)   & 0.337278 (0.121)        \\
 DWS  & 0.151183 (0.133)   & 0.234252 (0.068)         \\
 WS   &  0.153999 (0.365)  & 0.364165 (0.164)                 \\
 (WS/40)/(WS/48)   &  0.047009 (0.022)  & 0.151640 (0.004) \\
 \hline
 Sample Size & 142 & 376
\end{tabular}
\newline
\par
The graphs for the strongest correlation in both leagues (both the win shares statistic) are shown below:
\newline
\includegraphics[width=.85\textwidth]{NBA_WS_Graph.png}
\newline
\includegraphics[width=.85\textwidth]{WNBA_WS_Graph.png}

\par
Finally, within the Comp\_t\_tests.py Python script, analysis was conducted on the salaries of the top players in each league. Instead of looking at 
simple statistics, work that has been done by DiCicio \citep{Hoop_Dreams}, this method used the advanced statistics from the data set. Six statistics were selected: Minutes 
Played (MP), Player Efficiency Rating (PER), Free Throw rate (FTr), Offensive Win Shares (OWS), Defensive Win Shares (DWS), and Win Shares (WS). The 
script was written to sort the datasets by these statistics, and the top 10 players in each league in these statistics were extracted. Then the salaries 
were compared to each respective league's median salary, and the average of these top 10 players relations to the average were compiled. The entries in 
the graph can be interpreted as follows: If the value is greater than 1, than the top 10 players in that statistic are paid the value as a factor above 
league average. If the value is less than 1, than the top 10 players in that statistic are paid the value as a factor below league average. The chart is 
below:
\vspace*{1cm}
\begin{tabular}{*5l} 
\toprule
How Much Compared to Avg & the Top 10 Players Are Paid & For Advanced Stats \\\midrule

@@

\hspace*{10mm}
The data found from the initial t-test supported what had been found in the current literature. It was confirmed that in our 2018 data WNBA players are 
paid significantly less than NBA players, a known fact. It was also confirmed with a very high degree of certainty that WNBA players are paid less than 
NBA players proportional to league revenue (around 60 million for the WNBA, 8 billion for the NBA). This confirms the findings of David Berri \citep{WNBA_Revenue}, and was found 
to be particularly interesting while being generally intuitive based upon the narrative surrounding pay in women’s basketball.

@@

when compared to Win Shares is relatively high, the NBA graph looks much more traditionally distributed: Examining it finds the data generally following 

@@

\section{Conclusion}
\hspace*{10mm}
The motivation of this paper was to try and determine what goes into the pay gap in men’s and women’s basketball and primarily to see if the salary 
models were similar (just dramatically scaled down in the much poorer league). It was confirmed that the data for the 2018 season fit with precedent 
in terms of NBA salaries being much higher, both base and proportional to league revenue. The regression analysis revealed that the correlation 
coefficients for NBA salaries to selected impactful advanced statistics was in general much higher than that of the WNBA, revealing that the 
aforementioned salary models were dramatically different. It was, however, determined that finishing in the top 10 in said statistics resulted in pay 
increases over league median, however the NBA’s pay increases were significantly more due to their much larger spread in salaries. 
\par
Perhaps one of the most interesting conclusions is drawn when examining the graphs of NBA and WNBA pay distribution against Win Shares performance. 
Not only is there much less pay increase opportunity in the WNBA, but it seems as though almost all players are grouped into one of two pay rates, 
which may or may not be based solely upon performance.
\par
Next steps from this data involve researching what goes into league revenue and attempting to determine why it has never been profitable. 
It is important to acknowledge one last time that the WNBA is hamstrung by this lack of revenue, but the contract structures and much lower 
pay proportional to league revenue reinforce male dominant stereotypes in an already masculine world with a widespread gender pay gap in nearly
every industry. Ultimately, the NBA players get rewarded much more for performance than the WNBA, which may motivate them to improve more and bring 
more excitepment and popularity to the league. With less monetary motivation to perform, the WNBA continues to sell itself well short of its potential 
and does not give the players the respect that they deserve.
